---
title: "Intra-Session Reproducibility of Elbow Extension Velocity"
subtitle: "Statistical Analysis — ICC Study"
author: "Anaïs RAGON + Claude"
date: "12/05/2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

## 🔬 Scientific Context

> **Research question:** What is the reproducibility of instrumental angular velocity
> during passive elbow extension in healthy subjects?

> **Hypothesis:** Instrumental biomechanical parameters show good intra-session
> reproducibility in healthy subjects.

---

## 📋 Part 3 — Intra-Session Reproducibility Analysis

### Objective

Assess the intra-session reproducibility of angular velocity during **passive elbow
extension** across three speed conditions (**slow**, **medium**, **fast**) using the
Intraclass Correlation Coefficient (ICC).

### Inputs

- `velocity_trials_clean.csv` — exported from Part 2 (Python), one row per trial
- **Columns:** `patient_id` · `side` · `speed` · `trial` · `velocity` (°/s)

### Method

1. Data is reshaped from **long to wide format**: one row per subject, one column per trial
2. ICC is computed using the `irr` package:

| Parameter | Value | Meaning |
|-----------|-------|---------|
| `model` | `"twoway"` | Subjects **and** trials treated as random effects |
| `type` | `"agreement"` | Absolute agreement — penalises systematic differences |
| `unit` | `"single"` | Reliability of a single trial → **ICC(A,1)** |

3. Packages are installed only if missing (`requireNamespace`) — **safe on any machine**
4. Results interpreted using **Koo & Mae (2016)** thresholds:

| ICC value | Interpretation |
|-----------|---------------|
| < 0.50 | Poor |
| 0.50 – 0.75 | Moderate |
| 0.75 – 0.90 | Good |
| > 0.90 | Excellent |

---

## 📦 Packages

```{r packages}
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("irr",       quietly = TRUE)) install.packages("irr")
if (!requireNamespace("knitr",     quietly = TRUE)) install.packages("knitr")

library(tidyverse)
library(irr)
library(knitr)
```

---

## 📂 Load Data

```{r load_data}
# Finds the data folder relative to the .Rmd file location,
# regardless of the computer or working directory
data_path <- file.path(dirname(rstudioapi::getSourceEditorContext()$path),
                       "data", "velocity_trials_clean.csv")
df <- read.csv(data_path, stringsAsFactors = FALSE)

```

---

## 📐 ICC Analysis

```{r icc_function}
# Create unique subject ID (patient + side)
df <- df %>%
  mutate(subject = paste(patient_id, side, sep = "_"))

# Verify the column exists
cat("Columns available:", paste(names(df), collapse = ", "), "\n")
cat("Unique subjects:", paste(unique(df$subject), collapse = ", "), "\n")
```


```{r}
# ----------------------------------------------------------
# FUNCTION: compute ICC for one speed condition
# ----------------------------------------------------------
compute_icc <- function(speed_name, data) {

  df_wide <- data %>%
    filter(speed == speed_name) %>%
    select(subject, trial, velocity) %>%
    pivot_wider(names_from  = trial,
                values_from = velocity) %>%
    select(-subject)

  if (nrow(df_wide) < 2 || ncol(df_wide) < 2) {
    cat("\n[WARNING] Not enough data for speed:", speed_name, "\n")
    return(NULL)
  }

  res <- icc(df_wide,
             model = "twoway",
             type  = "agreement",
             unit  = "single")

  cat("\n============================\n")
  cat("Speed:", speed_name, "\n")
  print(res)
  return(res)
}
```

### Slow condition

```{r icc_slow}
icc_slow <- compute_icc("slow", df)
```

### Medium condition

```{r icc_medium}
icc_medium <- compute_icc("medium", df)
```

### Fast condition

```{r icc_fast}
icc_fast <- compute_icc("fast", df)
```

---

## 📊 Summary Table

```{r results_table}
extract_icc_row <- function(res, speed_name) {
  if (is.null(res)) return(NULL)
  data.frame(
    Speed          = speed_name,
    ICC            = round(res$value,   3),
    p_value        = round(res$p.value, 4),
    CI_lower       = round(res$lbound,  3),
    CI_upper       = round(res$ubound,  3),
    Interpretation = case_when(
      res$value < 0.50 ~ "Poor",
      res$value < 0.75 ~ "Moderate",
      res$value < 0.90 ~ "Good",
      TRUE             ~ "Excellent"
    )
  )
}

results_df <- bind_rows(
  extract_icc_row(icc_slow,   "Slow"),
  extract_icc_row(icc_medium, "Medium"),
  extract_icc_row(icc_fast,   "Fast")
)

kable(results_df,
      col.names = c("Speed", "ICC(A,1)", "p-value",
                    "95% CI lower", "95% CI upper", "Interpretation"),
      caption   = "Intra-session reproducibility — ICC results (Koo & Mae, 2016)")
```







