---
title: "Intra-Session Reproducibility of Instrumental Elbow Extension Velocity in Healthy Subjects: ICC Analysis"
author: "Anaïs RAGON"
date: "2026-05-12"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

---

## 0. Project Accessibility

The project is available on GitHub at this link: [https://github.com/Anais-rgn/Projet_R_Python_M2_REHAB_2026_Anais_Ragon](https://github.com/Anais-rgn/Projet_R_Python_M2_REHAB_2026_Anais_Ragon)

---

## 1. Scientific Background

### 1.1. Context

Spasticity is a frequent motor symptom following central nervous system injury (stroke, traumatic brain injury, spinal cord injury). It is defined as a **velocity-dependent increase in muscle tone**, causing functional limitations, pain, and significant impairment in daily life activities.

Current clinical assessment relies on standardised scales such as the **Modified Tardieu Scale (MTS)** and the Modified Ashworth Scale (MAS). However, these tools have well-documented limitations, in particular **poor inter- and intra-rater reproducibility**, which limits their reliability for longitudinal monitoring and clinical research.

Inertial Measurement Units (IMU) and portable dynamometers now offer the possibility of obtaining **objective, quantified, and reproducible measurements**. Combining these two technologies could provide a precise instrumental quantification of spasticity.

### 1.2. Aim

The aim of this pilot study is to assess the **intra-session reproducibility** of instrumental angular velocity during passive elbow extension in healthy subjects, as a preliminary step before the main clinical study on hemiparetic patients.

### 1.3. Method

Three healthy subjects performed passive elbow extension movements at **3 speed conditions**, each repeated **3 times**, on both arms (right and left), resulting in **6 limbs** analysed.

| Condition | Label | Target duration |
|-----------|-------|----------------|
| Slow | V1 — `slow` | ~5 s (metronome-guided) |
| Medium | V2 — `medium` | ~2 s |
| Fast | V3 — `fast` | < 1 s |

Instrumentation: **K-Push** (portable dynamometer, 1000 Hz) + **K-Move × 2** (inertial sensors / IMU, 250 Hz).

### 1.4. Participants

Three healthy subjects, bilateral measurements → **6 limbs** analysed (subjects: Ch, Lo, Ca).

### 1.5. Outcome Measure

The main outcome measure is **mean angular velocity** (°/s), computed for each extension trial as:

$$\text{velocity} = \frac{\theta_{max} - \theta_{min}}{t_{end} - t_{start}} \quad (°/s)$$

where $\theta_{max}$ and $\theta_{min}$ are the angle at the extension peak and the preceding flexion trough, respectively.

---

## 2. Aim of the Code

The objective of this R script is to assess the **intra-session reproducibility** of angular velocity measurements across three speed conditions (slow, medium, fast) using the **Intraclass Correlation Coefficient (ICC)**.

---

## 3. Data Organisation

**File:** `velocity_trials_clean.csv` — exported from the Python analysis (Part 2)

**Columns:**

| Column | Description |
|--------|-------------|
| `patient_id` | Subject identifier (e.g., Ch, Lo, Ca) |
| `side` | Limb side (`right` / `left`) |
| `speed` | Speed condition (`slow` / `medium` / `fast`) |
| `trial` | Trial number within each group (1, 2, 3) |
| `velocity` | Mean angular velocity (°/s) |

> **Note:** One row per trial. A unique subject identifier is created by combining `patient_id` and `side` (e.g., `Ch_right`).

---

## 4. Script Organisation

### 4.1. ICC Analysis by Speed Condition (section 5.3)

- **Aim:** Assess intra-session reproducibility of angular velocity for each speed condition
- **Input:** `velocity_trials_clean.csv` in `data/` folder
- **Calculation:** ICC(A,1) — twoway agreement model, single unit
- **Output:** ICC value, F-test (p-value), and 95% confidence interval per speed condition

### 4.2. Summary Table (section 5.4)

- **Aim:** Produce a formatted summary of all ICC results for comparison across speed conditions
- **Input:** ICC result objects (`icc_slow`, `icc_medium`, `icc_fast`)
- **Output:** Formatted table with ICC, p-value, 95% CI, and interpretation per speed

---

## 5. Code

---

### 5.1. Setup

These cells load the required libraries and the dataset, and ensure all columns are in the correct format for analysis.

```{r packages}
# Install packages only if not already present — safe on any machine
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("irr",       quietly = TRUE)) install.packages("irr")
if (!requireNamespace("knitr",     quietly = TRUE)) install.packages("knitr")

# Load libraries
library(tidyverse)
library(irr)
library(knitr)
```

```{r load_data}
# Load dataset exported from Python
df <- read.csv("data/velocity_trials_clean.csv", stringsAsFactors = FALSE)

# Create unique subject ID (patient + side)
df <- df %>%
  mutate(subject = paste(patient_id, side, sep = "_"))

# Display dataset dimensions and preview
cat("Rows:", nrow(df), "| Columns:", ncol(df), "\n")
cat("Unique subjects:", paste(unique(df$subject), collapse = ", "), "\n\n")

kable(head(df, 9), caption = "**Table 1.** Preview — velocity_trials_clean.csv")
```

---

### 5.2. Data Preparation for ICC

The ICC requires data in **wide format**: one row per subject, one column per trial.
This section reshapes the data separately for each speed condition.

```{r data_prep}
prepare_wide <- function(speed_name, data) {
  data %>%
    filter(speed == speed_name) %>%
    select(subject, trial, velocity) %>%
    pivot_wider(names_from = trial, values_from = velocity)
}

wide_slow   <- prepare_wide("slow",   df)
wide_medium <- prepare_wide("medium", df)
wide_fast   <- prepare_wide("fast",   df)

kable(wide_slow,   caption = "**Table 2.** Wide format — Slow condition (one column per trial)")
kable(wide_medium, caption = "**Table 3.** Wide format — Medium condition")
kable(wide_fast,   caption = "**Table 4.** Wide format — Fast condition")
```

---

### 5.3. Intraclass Correlation Coefficient (ICC)

The ICC is calculated to assess the **intra-session reproducibility** of angular velocity across the 3 trials for each speed condition.

The ICC(A,1) model is used with the following parameters:

| Parameter | Value | Meaning |
|-----------|-------|---------|
| `model` | `"twoway"` | Subjects **and** trials treated as random effects |
| `type` | `"agreement"` | Absolute agreement — penalises systematic differences |
| `unit` | `"single"` | Reliability of a single trial → **ICC(A,1)** |

> Reference: Koo & Mae, *J Chiropr Med*, 2016.

```{r icc_function}
compute_icc <- function(speed_name, data) {

  df_wide <- data %>%
    filter(speed == speed_name) %>%
    select(subject, trial, velocity) %>%
    pivot_wider(names_from  = trial,
                values_from = velocity) %>%
    select(-subject)    # keep numeric columns only

  if (nrow(df_wide) < 2 || ncol(df_wide) < 2) {
    cat("\n[WARNING] Not enough data for speed:", speed_name, "\n")
    return(NULL)
  }

  res <- icc(df_wide,
             model = "twoway",
             type  = "agreement",
             unit  = "single")

  cat("\n============================\n")
  cat("Speed:", speed_name, "\n")
  print(res)
  return(res)
}
```

#### 5.3.1. Slow condition

```{r icc_slow}
icc_slow <- compute_icc("slow", df)
```

#### 5.3.2. Medium condition

```{r icc_medium}
icc_medium <- compute_icc("medium", df)
```

#### 5.3.3. Fast condition

```{r icc_fast}
icc_fast <- compute_icc("fast", df)
```

---

### 5.4. Summary Table and Interpretation

This section produces a formatted summary of all ICC results.

```{r results_table}
extract_icc_row <- function(res, speed_name) {
  if (is.null(res)) return(NULL)
  data.frame(
    Speed          = speed_name,
    ICC            = round(res$value,   3),
    p_value        = round(res$p.value, 4),
    CI_lower       = round(res$lbound,  3),
    CI_upper       = round(res$ubound,  3),
    Interpretation = case_when(
      res$value < 0.50 ~ "Poor",
      res$value < 0.75 ~ "Moderate",
      res$value < 0.90 ~ "Good",
      TRUE             ~ "Excellent"
    )
  )
}

results_df <- bind_rows(
  extract_icc_row(icc_slow,   "Slow"),
  extract_icc_row(icc_medium, "Medium"),
  extract_icc_row(icc_fast,   "Fast")
)

kable(results_df,
      col.names = c("Speed", "ICC(A,1)", "p-value",
                    "95% CI lower", "95% CI upper", "Interpretation"),
      caption   = "**Table 5.** Intra-session reproducibility — ICC results (Koo & Mae, 2016)")
```

#### 5.4.1. Interpretation

The ICC ranges from 0 to 1 and measures the absolute agreement across repeated trials for the same subject.

The general interpretation scale of ICC values is as follows (Koo & Mae, 2016):

- **< 0.50:** Poor reliability
- **0.50 – 0.75:** Moderate reliability
- **0.75 – 0.90:** Good reliability
- **> 0.90:** Excellent reliability

```{r interpretation, echo=FALSE, results='asis'}
cat("> **Slow (ICC = 0.27, p = 0.067):** Poor reproducibility, non-significant.  \n")
cat("> The very wide confidence interval (–0.06 to 0.78) reflects high inter-trial variability,  \n")
cat("> likely due to the difficulty of maintaining a constant slow movement speed.\n\n")

cat("> **Medium (ICC = 0.11, p = 0.301):** Very poor reproducibility, non-significant.  \n")
cat("> The intermediate speed condition is the hardest to standardise,  \n")
cat("> resulting in maximum inter-trial variability.\n\n")

cat("> **Fast (ICC = 0.77, p < 0.001):** Good reproducibility, highly significant.  \n")
cat("> The confidence interval (0.38–0.96) is entirely positive,  \n")
cat("> suggesting that fast ballistic movements are more consistent across trials.\n\n")

cat("> ⚠️ **Critical limitation:** These results must be interpreted with major caution for two reasons:  \n")
cat("> (1) automatic trough detection was unreliable — several velocity values are incorrect;  \n")
cat("> (2) the sample size is very limited (n = 6 limbs), which explains the very wide confidence intervals  \n")
cat("> for slow and medium conditions. These ICC values do not carry valid clinical meaning.\n")
```

---

## 6. Conclusion

The analysis of intra-session reproducibility of angular velocity during passive elbow extension in healthy subjects reveals that:

- The **fast condition** shows good reproducibility (ICC = 0.77, p < 0.001), suggesting that ballistic fast movements are more stereotyped and therefore easier to measure consistently.

- The **slow and medium conditions** show poor and non-significant reproducibility (ICC = 0.27 and 0.11 respectively), with very wide confidence intervals that include zero or negative values.

However, these results must be interpreted with **major caution**:

1. **Unreliable trough detection:** The automatic detection of flexion minima failed in several cases, leading to incorrect velocity values.
2. **Insufficient sample size (n = 6 limbs):** Statistical power is very limited.

This pilot study validates the **end-to-end analysis pipeline** and identifies the key methodological challenge to address before the main clinical study: improving the automatic detection of movement onset.

---

## 7. References

Koo TK, Mae AY. A Guideline of Selecting and Reporting Intraclass Correlation Coefficients for Reliability Research. *J Chiropr Med.* 2016;15(2):155–163.

Shrout PE, Fleiss JL. Intraclass correlations: uses in assessing rater reliability. *Psychol Bull.* 1979;86(2):420–428.

Leuenberger K, Gonzenbach R, Wachter S, et al. A method to qualitatively assess arm use in stroke survivors at home. *Med Biol Eng Comput.* 2017;55:141–150.
